Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-23 manuscript/main.qmd
variability mid and high latitudes[@REFERENCE].
lewis adjustments main

```{=html}
<!-- discarded titles:
Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene
A Bayesian model of latitudinal temperature gradients applied to the Early Eocene
Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
Combining geochemical and ecological proxies...
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->
```
Accurately reconstructing large-scale palaeoclimate patterns from sparse local records is critical for understanding the evolution of Earth's climate. Particular challenges arise from the patchiness, uneven spatial distribution, and disparate nature of palaeoclimatic proxy records. Geochemical data typically provide temperature estimates via transfer functions derived from experiments. Similarly, transfer functions based on the climatic requirements of modern taxa exist for some fossil groups, such as pollen assemblages. In contrast, most ecological and lithological data (e.g. coral reefs and evaporites) only convey information on broad climatic requirements. Historically, most large-scale proxy-based reconstructions have used either geochemical or ecological data, but few studies have combined multiple proxy types into a single quantitative reconstruction. Large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into account the spatial distribution of samples, leading to biased temperature reconstructions. Here, we present a Bayesian hierarchical model to integrate ecological data with established geochemical proxies into a unified quantitative framework, bridging gaps in the latitudinal coverage of proxy data. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal sea surface temperature gradient and global average temperatures of the EECO. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data. We show that integrating multiple types of proxy data, and adequate prior information, has the potential to enhance quantiative palaeoclimate reconstructions, allowing for more accurate temperature estimates from  data with incomplete spatial sampling.

@@

Palaeoclimate, latitudinal temperature gradients, temperature proxies, Eocene, spatial bias, Bayesian

@@

Understanding the long-term evolution of Earth's climate system and contextualising current global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018; @Tierney2020]. Recent advances in the synthesis of palaeoclimate data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, yet a fundamental challenge remains: the proxy record of past climates is spatially incomplete and afflicted by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Acknowledging the assumptions and limitations inherent in geochemical temperature proxies, e.g. experimentally derived calibrations, influences from seasonality, dissolution effects and differential preservation [e.g. @Tierney2017], can allow for robust estimates of palaeotemperatures at a local scale. However, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. Principally, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of other regions (e.g. a high proportion of samples from tropical areas). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions [e.g. @song2019]. However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, which requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data. Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

Spatial gaps in the palaeoclimate record can also be addressed through the integration of additional data. For example, lithological and fossil data can be used to infer past climatic conditions based on analogous modern sediments [@Chandra2021], or based on the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Despite this potential, the integration of geochemical proxy data with other sources of information (e.g. ecological data) has rarely been realised in a rigorous, quantitative framework [@Burgener2023].

Here, we present a novel Bayesian hierarchical model [see e.g. @Gelman2013; @Mcelreath2018] that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g. @Weitzel2019; @Yang2022; @Burgener2023]. This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework. We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage. This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g. @Bijl2009; @Keating-Bitoni2011], lack. We test the robustness of this method using down-sampled, simulated temperature gradients.

@@

::: {#table-1}
Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies. The gradient values denote the SST difference between the equator and the polar circle, or other types of gradients. For comparison, a gradient derived from an atmosphere-ocean general circulation model (GCM) ensemble is also shown.

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~. For our analyses, this dataset was restricted to the EECO (defined as 53.8--49.1 Ma) and samples originating from near the ocean surface or mixed layer. Samples labelled as "thermocline" or "sub-thermocline" were excluded. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained (Fig. 1). For a detailed description of each proxy see @Hollis2019.

@@

The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 plate rotation model via the palaeoverse R package [version 1.2.0, @jones2023]. The midpoint age of the EECO (51.2 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeogeographic reconstruction.

@@

We use this generalised logistic function because it can follow the equatorial and polar asymptotes observed in the modern, latitudinal SST gradient, but can also accommodate a variety of other shapes, while consisting of only four shape parameters. This flexibility is primarily achieved by shifting the location of the curve along the latitudinal axis by varying $M$, and by altering the steepness of the curve by varying $B$. For example, one limb of a second-order polynomial as in @Bijl2009 can be approximated by increasing $M$ towards high latitudes, and decreasing $B$ to reduce the steepness of the curve. The model is designed for modelling the average gradient across both hemispheres, but can also be applied to individual hemispheres, to assess hemispherical differences (Supplementary Material).

@@

We show an expanded model that includes uncertainties on individual temperature observations in the Supplementary Material.

@@

**A.** Predicted seawater surface temperatures are not allowed to be $<< -2^\circ C$, the freezing point of sea water. The highest prior density of $A$ is placed around $0^\circ C$, and it slowly tapers off towards higher temperatures. This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as $$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$${#eq-A}
where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we used the empirical, modern gradient, representative of an icehouse climate, and generated three idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse (modern), greenhouse, and extreme greenhouse [@Frakes1992]. The idealised gradients serve to test whether our model setup is able to infer gradients that are strongly different from the modern from a varying amount of samples.

We created test data from these gradients as follows: We randomly sampled (1000 iterations) latitudes at sample sizes of 5, 10 and 20, with the probability of a latitude being sampled scaling with the decreasing surface area towards higher latitudes; i.e. lower latitudes are sampled more frequently. For the largest sample size (n = 34), we used the latitudes of the EECO data set of this study in all iterations. For each latitude, we inferred a location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 3.8, which is the average standard deviation of the EECO geochemical proxy data. With that, we aim to simulate randomly distributed errors in the proxy data, which could arise from miscalibrations, measurement errors, seasonal effects, ect. We acknowledge that this approach cannot quantify the potential impact of systematic offsets that may bias all proxy data in the same direction, nor do we know whether a standard deviation of 3.8 is actual, average magnitude of uncertainty that the proxy compilation is afflicted with.

@@

To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to annual sea surface mean temperatures from Bio-Oracle [@Assis2018], aggregated to a $1^\circ \times 1 ^\circ$ raster (*n* = 46,131). The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modelled gradient and the empirical temperature averages in 1&deg; latitude bins. Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.

@@

We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters [see @Gilks1995; @Gelman2013]. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in. Every 10th iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation. The re-sampled, simulated gradients were modelled in one chain with 10,000 iterations for each of the 1,000 random samples. 5,000 iterations each were discarded as burn-in, and every 25th iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs. For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in. For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every 100th iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Fig. S1).

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme icehouse, icehouse, greenhouse, and extreme greenhouse. The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column). To generate the random samples, noise from a normal distribution with a standard deviation of 3.8 was added to all temperatures. The blue shadings depict the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modelled gradient. The median (50%) $R^2$ value along with the 95% credible intervals from all model runs are shown. Each gradient is depicted in absolute latitude.
:::

@@

Our Bayesian model is able to accurately model a range of idealised temperature gradients, ranging from extreme icehouse to 'super greenhouse' scenarios (Fig. 2). Random latitudinal sampling results in highly accurate reconstructions at a sample sizes as low as 10 for the icehouse scenarios (95 % CI of $R^2$ > 0.9). Greenhouse scenarios require additional samples to accurately predict high-latitude temperatures. This is because in the absence of high-latitude samples, the modelled gradient is heavily influenced by the priors, which we based on the modern, the only empirically known latitudinal temperature gradient. A sampling distribution resembling that of the early Eocene data set used in this study allows for a highly accurate reconstruction of even the extreme greenhouse scenario (95 % CI of $R^2$ > 0.95).

The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude (black line) is explained by the modelled gradient ($`r 100*as.numeric(R2_modern_median)` \%$). The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only slightly higher at `r modern_gradient$"50%"`&deg;C. The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, very similar to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).

@@

The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig. 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"`&deg;C (95% CI: `r tdiff_lat0$"2.5%"`--`r tdiff_lat0$"97.5%"`&deg;C) higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"`&deg;C (`r tdiff_lat90$"2.5%"`--`r tdiff_lat90$"97.5%"`&deg;C) higher. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`&deg;C (`r eocene_gradient$"2.5%"`--`r eocene_gradient$"97.5%"`&deg;C) for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg;), which is a lot lower at `r eocene_gradient_polar_circle$"50%"`&deg;C (`r eocene_gradient_polar_circle$"2.5%"`--`r eocene_gradient_polar_circle$"97.5%"`&deg;C).

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient--`r eeco_sdy$"50%"` (`r eeco_sdy$"2.5%"`--`r eeco_sdy$"97.5%"`)--which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`. This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

@@

Estimates of the median, latitudinal sea surface temperature gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model. The purple ribbon (shading) depict the 95% credible interval of the Eocene gradient, the uncertainty of the modern gradient is too low to be visible. Points within the plot depict the geochemical (e.g. TEX~86~) and ecological (e.g. mangroves) data. Geochemical data are plotted by their point estimate temperature value. Ecological data are plotted at the mean temperature values of their respective normal distributions.
:::

@@

The authors are grateful to all those who have enabled this work by collecting, measuring, collating and screening geochemical and fossil data. The contribution of L.A.J. was supported by a Juan de la Cierva-formación 2021 fellowship (FJC2021-046695-I/MCIN/AEI/10.13039/501100011033) from the European Union "NextGenerationEU"/PRTR. For the purpose of open access, the authors have applied a Creative Commons Attribution (CC BY) licence to any Author Accepted Manuscript version arising from this submission.
<!-- regarding funding acknowledgement, I talked to my boss and he says I don't need to credit my funding as I have done this work largely in my spare time -->

>>>>

```{=html}
<!-- discarded titles:
Bayesian multi-proxy reconstruction of latitudinal temperature gradients applied to the Early Eocene
A Bayesian model of latitudinal temperature gradients applied to the Early Eocene
Bayesian multi-proxy reconstruction of latitudinal temperature gradients for the Early Eocene
Bayesian multi-proxy reconstruction of the early Eocene latitudinal temperature gradient
Combining geochemical and ecological proxies...
Modelling latitudinal temperature gradients from geological proxies: An early Eocene example
"Improving palaeoclimatic reconstructions with ecological data: How hot was the early Eocene?"
"Integrating ecological and geochemical proxies for palaeoclimatic reconstruction: An early Eocene example"
"The Early Eocene Climate Optimum reconstructed with Bayesian multi-proxy integration"
"An Early Eocene temperature gradient modelled with geological and ecological proxies"
-->
```
Accurately reconstructing large-scale palaeoclimatic patterns from sparse local records is critical for understanding the evolution of Earth's climate. Particular challenges arise from the patchiness, uneven spatial distribution, and disparate nature of palaeoclimatic proxy records. Geochemical data typically provide temperature estimates via transfer functions derived from experiments. Similarly, transfer functions based on the climatic requirements of modern taxa exist for some fossil groups, such as pollen assemblages. In contrast, most ecological and lithological data (e.g. coral reefs and evaporites) only convey information on broad climatic requirements. Historically, most large-scale proxy-based reconstructions have used either geochemical or ecological data, but few studies have combined multiple proxy types into a single quantitative reconstruction. Large spatial gaps in existing proxy records have often been bridged by simple averaging, without taking into account the spatial distribution of samples, leading to biased temperature reconstructions. Here, we present a Bayesian hierarchical model to integrate ecological data with established geochemical proxies into a unified quantitative framework, bridging gaps in the latitudinal coverage of proxy data. We apply this approach to the early Eocene climatic optimum (EECO), the interval with the warmest sustained temperatures of the Cenozoic. Assuming the conservation of thermal tolerances of modern coral reefs and mangrove taxa, we establish broad sea surface temperature ranges for EECO coral reef and mangrove sites. We integrate these temperature estimates with the EECO geochemical shallow marine proxy record to model the latitudinal sea surface temperature gradient and global average temperatures of the EECO. Our results confirm the presence of a flattened latitudinal temperature gradient and unusually high polar temperatures during the EECO, which is supported by high-latitude ecological data. We show that integrating multiple types of proxy data, and adequate prior information, has the potential to enhance quantitative palaeoclimatic reconstructions, improving temperature estimates from datasets with limited spatial sampling.

@@

Palaeoclimate, latitudinal temperature gradients, temperature proxies, Eocene, sampling bias, Bayesian

@@

Understanding the long-term evolution of Earth's climate system and contextualising contemporary global warming relies on accurate reconstructions of past climates [@Royer2004; @Burke2018; @Tierney2020]. Recent advances in the synthesis of palaeoclimatic data [e.g. @Veizer2015; @Hollis2019; @song2019; @Grossman2022; @Judd2022] are offering unprecedented insights into the complex and dynamic nature of the Earth's climate system, yet a fundamental challenge remains: the proxy record of past climates is spatially incomplete and afflicted by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Acknowledging the assumptions and limitations inherent in geochemical temperature proxies, such as experimentally derived calibrations, influences from seasonality, dissolution effects and differential preservation [e.g. @Tierney2017], can enable robust estimates of palaeotemperature at local scales. However, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. Principally, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of other regions (e.g. a high proportion of samples from tropical areas). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions [e.g. @song2019]. However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, which requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [e.g. @Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimatic data [e.g. @Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data. Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

Spatial gaps in the palaeoclimatic record can also be addressed through the integration of additional data. For example, lithological and fossil data can be used to infer past climatic conditions based on analogous modern sediments [@Chandra2021], or based on the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Despite this potential, the integration of geochemical proxy data with other sources of information (e.g. ecological data) has rarely been realised in a rigorous, quantitative framework [@Burgener2023].

Here, we present a novel Bayesian hierarchical model [e.g. @Gelman2013; @Mcelreath2018] that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled palaeoclimatic data. The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g. @Weitzel2019; @Yang2022; @Burgener2023]. This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological palaeoclimatic proxies in a common, quantitative framework. We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage. This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g. @Bijl2009; @Keating-Bitonti2011], lack. We test the robustness of this method using down-sampled, simulated latitudinal temperature gradients.

@@

::: {#table-1}
Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies. The gradient values denote the SST difference between the equator and the polar circle, or other types of gradients. For comparison, a gradient derived from an atmosphere-ocean general circulation model (GCM) ensemble, and a range of gradients from a model intercomparison project, are also shown.

@@

Geochemical climatic proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~. For our analyses, this dataset was restricted to the EECO (defined as 53.8--49.1 Ma) and samples originating from near the ocean surface or mixed layer. Consequently, samples labelled as "thermocline" or "sub-thermocline" were excluded. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained (Fig. 1). For a detailed description of each proxy see @Hollis2019.

@@

The palaeogeographic distribution of geochemical and ecological data was reconstructed using the @merdith2021 Global Plate Model via the palaeoverse R package [version 1.2.0, @jones2023]. The midpoint age of the EECO (51.2 Ma), along with the present-day coordinates of geochemical and ecological data, were used for palaeogeographic reconstruction.

@@

We use this generalised logistic function because it can follow the equatorial and polar asymptotes observed in the modern, latitudinal SST gradient, but can also accommodate a variety of other shapes, while consisting of only four shape parameters. This flexibility is primarily achieved by shifting the location of the curve along the latitudinal axis by varying $M$, and by altering the steepness of the curve by varying $B$. For example, one limb of a second-order polynomial as in @Bijl2009 can be approximated by increasing $M$ towards high latitudes, and decreasing $B$ to reduce the steepness of the curve. The model is designed for modelling the average gradient across both hemispheres, but can also be applied to individual hemispheres, to assess hemispherical differences (see Fig. S4).

@@

We show an expanded model that includes uncertainties on individual temperature observations in the Supplementary Material (Fig. S5).

@@

**A.** Predicted seawater surface temperatures are not allowed to be $<< -2$&deg;C, the freezing point of sea water. The highest prior density of $A$ is placed around 0&deg;C, and it slowly tapers off towards higher temperatures. This shape is achieved by placing a skew-normal prior on the lower asymptote, specified as $$A \sim SN(\xi=-3.0,\omega=12,\alpha_{SN}=30),$${#eq-A}
where $\xi$, $\omega$, and $\alpha_{SN}$ are the location, scale and shape parameters.

@@

To test whether our logistic regression model can adequately describe different latitudinal temperature gradients at various sample sizes, we used the empirical, modern gradient, representative of an icehouse climate, and generated three idealised gradients that emulate potential climatic states throughout Earth's geological history: extreme icehouse, icehouse (modern), greenhouse, and extreme greenhouse [@Frakes1992]. The idealised gradients serve to test whether our model setup is able to infer gradients that are strongly different from the modern from a varying number of samples.

We created test data from these gradients as follows: We randomly sampled (1000 iterations) latitudes at sample sizes of 5, 10 and 20, with the probability of a latitude being sampled scaling with the decreasing surface area towards higher latitudes, i.e. lower latitudes are sampled more frequently. For the largest sample size (n = 34), we used the latitudes of the EECO data set of this study in all iterations. For each latitude, we inferred a location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 3.8, which is the average standard deviation of the EECO geochemical proxy data. With that, we aim to simulate randomly distributed errors in the proxy data, which could arise from miscalibrations, measurement errors, seasonal effects, *ect*. We acknowledge that this approach cannot quantify the potential impact of systematic offsets that may bias all proxy data in the same direction, nor do we know whether a standard deviation of 3.8 is actual, average magnitude of uncertainty that the proxy compilation is afflicted with.

@@

To test whether our model can accurately depict the shape of the modern sea surface temperature gradient, and to facilitate comparison with the Eocene gradient, we applied our model to mean annual sea surface temperatures from Bio-Oracle [@Assis2018], aggregated to a spatial grid resolution of $1^\circ \times 1 ^\circ$ (*n* = 46,131). The $R^2$ for the modern gradient was calculated as above [@Gelman2019], comparing the modelled gradient and the empirical temperature averages in 1&deg; latitude bins. Only the medians are reported for the modern gradient, as the 95% credible intervals are extremely narrow due to the high precision of the posterior estimates.

@@

We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters [see @Gilks1995; @Gelman2013]. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in. Every $10^{th}$ iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation. The re-sampled, simulated gradients were modelled in one chain with 10,000 iterations for each of the 1,000 random samples. 5,000 iterations each were discarded as burn-in, and every $25^{th}$ iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs. For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in. For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every $100^{th}$ iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Fig. S1).

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples). Each row depicts a different latitudinal temperature gradient that represents idealised or observed climatic states: idealsed extreme icehouse, greenhouse, and extreme greenhouse gradients, and the modern gradient, which represents an icehouse state. The black line illustrates the original gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column). To generate the random samples, noise from a normal distribution with a standard deviation of 3.8 was added to all temperatures. The blue shadings depict the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modelled gradient. The median (50%) $R^2$ value along with the 95% credible intervals from all model runs are shown. Each gradient is depicted in absolute latitude.
:::

@@

Our Bayesian model is able to model a range of idealised temperature gradients, ranging from extreme icehouse to 'super greenhouse' scenarios (Fig. 2). Random latitudinal sampling results in accurate reconstructions for most random samples at sample sizes of 10 and 20 for the icehouse scenarios (median of $R^2$ > 0.9). Greenhouse scenarios perform somewhat worse due to the increased uncertainty at high latitudes (median of $R^2$ > 0.7 at sample sizes 10 and 20). A sampling distribution resembling that of the early Eocene dataset used in this study allows for accurate reconstruction of all scenarios, although the $R^2$ is still relatively low in the extreme greenhouse scenario, as a perfectly flat gradient prediction by the model would result in an $R^2$ of 0, despite the original gradient being very flat. This also explains the low lower bounds of the 95% credible intervals in the greenhouse scenarios.

The average, modern temperature gradient can be closely approximated with our model when using the full modern SST dataset (Fig. 3); almost all of the variation in the empirical median temperatures in bins of 1&deg; absolute latitude (black line) is explained by the modelled gradient ($`r 100*as.numeric(R2_modern_median)` \%$). The empirical gradient spans `r modern_gradient_empirical`&deg;C from the equator to the poles, the modelled gradient is only slightly higher at `r modern_gradient$"50%"`&deg;C. The modern, global mean temperature (GMST) based on our modelled, median gradient is `r motemp_global$"50%"`&deg;C, which is nearly equal to the GMST derived from the empirical median gradient (`r motemp_empirical_global`&deg;C).

@@

The Eocene temperature gradient reconstructed with our Bayesian model is starkly different from the modern (Fig. 4). Modelled, median equatorial temperatures are `r tdiff_lat0$"50%"`&deg;C (95% CI: `r tdiff_lat0$"2.5%"`--`r tdiff_lat0$"97.5%"`&deg;C) higher for the EECO, and polar temperatures are `r tdiff_lat90$"50%"`&deg;C (`r tdiff_lat90$"2.5%"`--`r tdiff_lat90$"97.5%"`&deg;C) higher. This results in a flattened latitudinal temperature gradient of `r eocene_gradient$"50%"`&deg;C (`r eocene_gradient$"2.5%"`--`r eocene_gradient$"97.5%"`&deg;C) for the EECO, as opposed to `r modern_gradient$"50%"`&deg;C for the modern. To facilitate the comparison with latitudinal gradients reported in the literature, which sometimes do not report temperatures at very high latitudes, we report also the EECO gradient between the equator and the modern-day polar circle (66.6&deg; latitude), which is markedly lower at `r eocene_gradient_polar_circle$"50%"`&deg;C (`r eocene_gradient_polar_circle$"2.5%"`--`r eocene_gradient_polar_circle$"97.5%"`&deg;C).

The high variability of EECO palaeotemperature proxies, particularly in the mid-latitudes, and the scarcity of high-latitude data, result in substantial uncertainties in the modelled temperature gradient. This is reflected in the residual standard deviation ($\sigma$) of the EECO gradient--`r eeco_sdy$"50%"`&deg;C (`r eeco_sdy$"2.5%"`--`r eeco_sdy$"97.5%"`&deg;C)--which is more than double the  $\sigma$ for the modern gradient, `r modern_sdy$"50%"`&deg;C. This is illustrated by the drastic departure of some of the proxy data from the gradient estimates (Fig. 4).

@@

Estimates of the median, latitudinal sea surface temperature gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model. The purple ribbon (shading) depicts the 95% credible interval of the Eocene gradient, the uncertainty of the modern gradient is too low to be visible. Points within the plot depict the geochemical (e.g. TEX~86~) and ecological (e.g. mangroves) data. Geochemical data are plotted by their point estimate temperature value. Ecological data are plotted at the mean temperature values of their respective normal distributions.
:::

@@

The authors are grateful to all those who have enabled this work by collecting, measuring, collating, and screening geochemical and fossil data. The contribution of L.A.J. was supported by a Juan de la Cierva-formación 2021 fellowship (FJC2021-046695-I/MCIN/AEI/10.13039/501100011033) from the European Union "NextGenerationEU"/PRTR. For the purpose of open access, the authors have applied a Creative Commons Attribution (CC BY) licence to any Author Accepted Manuscript version arising from this submission.
<!-- regarding funding acknowledgement, I talked to my boss and he says I don't need to credit my funding as I have done this work largely in my spare time -->
